Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-21 rnni_convexity.tex
proof caterpillar convexity

\begin{proof}
We prove this Theorem by backwards induction on the caterpillar distance $l = d_c(T,R)$ between caterpillar trees $T$ and $R$.
For this backwards induction the maximum caterpillar distance, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}), is the base case.

Induction basis: $l = \frac{(n-1)(n-2)}{2}$\\
With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$.
This means that there is a shortest path between $T$ and $R$ that is a caterpillar path.

@@

Since $l < \frac{(n-1)(n-2)}{2}$, there must be one of the loops starting in Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort} ($\csort$) where not every considered pair of taxa swaps positions.
Let us assume that loop number $k$ is the first one where the number of swaps is not maximal.
It might well be that this happens in the first loop $k=1$ already.
Since we assume that the number of swaps in loop $k$ is not maximal, there is a pair $a,b$ of elements that is compared in the algorithm but does not swap positions.
And because we consider the first time that this happens, $a$ is either one of the first two elements in the sequence, or it previously swapped with an element $c$ that has been between $a$ and $b$ before loop number $k$.

Let us first consider the latter case.
The sequence $L_{T'}$ at the time where $a$ and $b$ are compared but not swapped, looks as follows:

\vspace{12pt}
\begin{tabular}{c|c|c|c|c|c|c|c|c|}
\cline{2-9}
$L_{T'}$: & $\ldots$ & $c$ & $a$ & $b$ & $\ldots$& $x_{k-1}$ & $\ldots$ & $x_1$\\
\cline{2-9}
\end{tabular}
\vspace{12pt}

Note that in this illustration of $L_{T'}$ the elements $x_1, \ldots, x_{k-1}$ are in the same place as they are in $L_R$.
However, it could be that $k=1$, which means that there are no such elements at the end of $L_{T'}$ that are already sorted the same way as in the destination tree.

The last move before comparing $a$ and $b$ is a swap of $a$ with an element $c$.
Since $a$ and $c$ swapped, but $a$ and $b$ did not, we can follow that $b$ and $c$ appear in $L_R$ in the same order as they do in $L_{T'}$.
Furthermore, $c$ and $b$ must have been neighbours in $L_T$, because we assume that the comparison of $a$ and $b$ is the first one where the compared elements are not swapped in the sequence.
Let now $L_{\hat T}$ be a list that equals $L_T$ with the only exception that $b$ and $c$ are swapped.
We can follow from our observations above, that the distance between $T$ and $\hat T$ is $d(T, \hat T) = 1$ and that $\csort$ needs one move more for a path from $\hat T$ to $R$ than from $T$ to $R$.
This results in $d_c(\hat T, R) = d(T,R) + 1$.

@@

Let us now consider the other case:
of the two elements $a$ and $c$, that are compared but not swapped, one, say $a$, is either the first or the second one in $L_{T'}$.
Then the list looks like this:

\vspace{12pt}
\begin{tabular}{c|c|c|c|c|c|c|c|c|}
\cline{2-8}
$L_{T'}$: & $a$ & $b$ & $c$ & $\ldots$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
\cline{2-8}
\end{tabular}
\vspace{12pt}

If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with caterpillar distance $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
In the list $L_T$ representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.
This is due to the fact that we are currently considering the first comparison of elements where there is no swap.
Without loss of generality we assume that $a$ is neighbour of $c$ in $L_T$.
Then an exchange of $a$ and $c$ in $L_T$ yields a list $L_{\hat T}$ representing a tree $\hat T$ with the desired properties.
As in the previous case we can then apply the induction hypothesis on $\hat T$ and conclude that there is a shortest path between $T$ and $R$ that only consists of caterpillar trees.
\end{proof}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{NNI_NP_proof}
\vspace{12pt}
\caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
All leaves labelled with dots represent taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs. squares) shared between $T$ and $R$ are preserved.
Instead, the depicted path where $n$ cherries are built and then sorted before being resolved is a shortest path}
\label{fig:NNI_NP_proof}
\end{figure}

@@

\section{Ideas}

[This section includes some ideas that we could possibly include in the paper]

Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.

\begin{proposition}
If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
\label{proposition:lower_bound_distance}
\end{proposition}

The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
$\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.

\begin{conjecture}
Let $T$ and $R$ be ranked trees.
The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
\end{conjecture}

\begin{figure}[H]
\centering
\includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
\vspace{12pt}
\caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
\label{fig:findpath_not_symmetric}
\end{figure}

% idea for description of max distance caterpillar trees:
\begin{lemma}
Let $T$ and $R$ be caterpillar trees.
It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
\todo{define induced triplets}
\end{lemma}

This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)

\begin{proof}
%TODO
% Insert the proof here
Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
\end{proof}

According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.

\subsection{Partition lattice}

% Define Partition Lattice $\Pi_n$ and max chains in that lattice!

\begin{theorem}
The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
The corresponding metric spaces are isometric.
\end{theorem}

>>>>

\begin{proof}
We prove this Theorem by backwards induction on the caterpillar distance $l = d_c(T,R)$ between caterpillar trees $T$ and $R$.
For this backwards induction the maximum caterpillar distance $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:caterpillar_dist=diameter}) is the base case.

Induction basis: $l = \frac{(n-1)(n-2)}{2}$\\
With Lemma~\ref{lemma:caterpillar_dist=diameter} one can see that for each pair of caterpillar trees $T$, $R$ with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$, it is $d(T,R) = d_c(T,R)$.
Thus there is a shortest path between $T$ and $R$ that is a caterpillar path.

@@

Since $l < \frac{(n-1)(n-2)}{2}$, there must be a loop in $\csort$ (Line~\ref{alg:csort:line:loop} of Algorithm~\ref{alg:csort}) where not every pair of taxa that is compared exchanges.
Let us consider the first pair of taxa $a,b$ that is compared but does not swap positions, where $a$ is below $b$.
Since the first time that this happens, $a$ is either one of the first two elements in the sequence, or it previously swapped with an element $c$ that has been between $a$ and $b$ before.

We start with considering the latter case.
The list representation of the tree at the point where $a$ and $b$ are compared but not swapped, looks as follows:
\[[\ldots, c, a, b, \ldots, x_k, \ldots, x_1]\]

Note that here we assume that the elements $x_1, \ldots, x_k$ are in the same place as they are in the destination tree $R$.
It could be that $k=1$, which means that there are no such elements.

Since $a$ and $c$ swap, but $a$ and $b$ do not, we can follow that $b$ and $c$ appear in $R$ in the same order as they do in the current tree.
Furthermore, $c$ and $b$ must have been neighbours in $T$, because we assume that the comparison of $a$ and $b$ is the first one where the compared elements are not swapped in the sequence.
Let now $\hat T$ be a tree that equals $T$ with the only exception that $b$ and $c$ are exchanged.
We can follow from our observations above that the distance between $T$ and $\hat T$ is $d(T, \hat T) = 1$ and that $\csort$ needs one move more for a path from $\hat T$ to $R$ than from $T$ to $R$.
This results in $d_c(\hat T, R) = d(T,R) + 1$.

@@

Let us now consider the other case:
of the two taxa that are compared but do not swap positions, one is part of the cherry current tree.
Let $a$ and $b$ be the cherry taxa and $c$ their neighbour such that neither $a$ nor $b$ echanges with $c$.
Then the list looks like this:
\[[a, b, c, \ldots, x_k, \ldots, x_1]\]

If there is no swap of either $a$ or $b$ with $c$, we can construct a tree $\hat T$ with caterpillar distance $d_c(\hat T, R) = l+1$ and $d(T,\hat T) = 1$ in a similar way as in the previous case.
In the list representing the start tree $T$, either $a$ or $b$ must be neighbour of $c$.
This is due to the fact that we are currently considering the first comparison of elements where there is no swap.
We can assume without loss of generality that $a$ is neighbour of $c$ in $T$.
Then an exchange of $a$ and $c$ in $T$ yields a tree $\hat T$ with the desired properties.
As in the previous case we can then apply the induction hypothesis on $\hat T$ and conclude that there is a shortest path between $T$ and $R$ that only consists of caterpillar trees.
\end{proof}

@@

\begin{figure}[H]
\centering
\includegraphics[width=.8\textwidth]{NNI_NP_proof}
\vspace{12pt}
\caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
All leaves labelled with dots represent taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters (dots vs. squares) shared between $T$ and $R$ are preserved.
Instead, the depicted path where $n$ cherries are built and then sorted before being resolved is a shortest path}
\label{fig:NNI_NP_proof}
\end{figure}

@@

% \section{Ideas}
%
% [This section includes some ideas that we could possibly include in the paper]
%
% Using the same argument as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.
%
% \begin{proposition}
% If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
% \label{proposition:lower_bound_distance}
% \end{proposition}
%
%
% The paths computed by $\findpath$ depend on the order of input trees $T$ and $R$:
% $\findpath(T,R)$ does not necessarily produce the reverse path of $\findpath(R,T)$.
% An example can be found in Figure~\ref{fig:findpath_not_symmetric}.
% However, simulations suggest that the lengths of the two paths computed by $\findpath$ for one pair of trees are equal.
%
% \begin{conjecture}
% Let $T$ and $R$ be ranked trees.
% The lengths of paths $\findpath(T,R)$ and $\findpath(R,T)$ are equal.
% \end{conjecture}
%
% \begin{figure}[H]
% \centering
% \includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
% \vspace{12pt}
% \caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
% \label{fig:findpath_not_symmetric}
% \end{figure}
%
% % idea for description of max distance caterpillar trees:
% \begin{lemma}
% Let $T$ and $R$ be caterpillar trees.
% It is $d_c(T,R) = \Delta(\rnni)$ if, and only if, $T$ and $R$ share no induced triplet.
% \todo{define induced triplets}
% \end{lemma}
%
% This Lemma does only hold for caterpillar trees and not for trees with more than one cherry.
% Furthermore, triplets do not represent ranked trees uniquely as they do not contain information about the ranks of internal nodes (but they can uniquely define caterpillar trees as these only have one cherry)
%
% \begin{proof}
% %TODO
% % Insert the proof here
% Fix $T = (\ldots(1,2) \ldots ,n)$, use induction on $n$ and $\csort$ (taxon $n$ must be in cherry of $R$ if distance is max).
% \end{proof}
%
% According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given caterpillar tree $T$ is $(n-1)!$.
% The number of caterpillar trees with distance $\Delta(\rnni)$ from a given caterpillar tree is $2^{n-2}$.
%
%
% \subsection{Partition lattice}
%
% % Define Partition Lattice $\Pi_n$ and max chains in that lattice!
%
% \begin{theorem}
% The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
% The corresponding metric spaces are isometric.
% \end{theorem}

\section{Discussion}
